Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-26 paper/main.tex
minor things in discussion

In this work, we introduce a novel Bayesian hierarchical framework that leverages deep 
Gaussian processes (DGPs) to model smooth latent functions from correlated functional outputs, 
such as those arising in cosmological simulations. Our model, DGP.FCO, builds on the compositional nature 
of DGPs and introduces a third layer to explicitly model correlated observational errors within 
each spectrum. Through a simulation study, we demonstrate that DGP.FCO offers improved uncertainty 
quantification compared to existing methods.

We apply DGP.FCO to both the CAMB and Mira-Titan datasets, generating posterior mean estimates 
and predicting at held-out cosmological settings. Across both datasets, DGP.FCO 
consistently performs well against competitive baselines, showcasing strong generalization and 
flexibility in complex, functional data regimes.

Looking ahead, several avenues for future research arise. Beyond modeling dark matter power 
spectra, DGP.FCO could be extended to hydrodynamical simulation outputs in cosmology, which 
often share similar functional structures and error correlations. More broadly, this approach may be 
applicable in other scientific domains involving structured functional outputs, especially those 
with spatial or temporal dependencies.

In addition, generating predictions over large cosmological grids opens the door to detailed 
sensitivity analyses, enabling researchers to disentangle parameter interactions and identify 
the most influential physical drivers of the spectra. From a modeling standpoint, further 
developments could include joint modeling of latent layers across cosmologies, and extensions 
to redshifts beyond $z=0$. Lastly, a hybrid framework combining DGP.FCO with Cosmic Emu, leveraging 
the strengths of each at different scales of $k$, may yield even more accurate predictions.

>>>>

In this work, we introduced a novel Bayesian hierarchical framework that leverages deep 
Gaussian processes (DGPs) to model smooth latent functions from correlated functional outputs, 
such as those arising in cosmological simulations. Our model (DGP.FCO) builds on the 
compositional nature of DGPs and introduces an additional outer layer to explicitly 
model correlated observational errors within each spectrum. Through a simulation study, 
we demonstrated that DGP.FCO offers improved uncertainty 
quantification compared to existing methods.

We applied DGP.FCO to both the CAMB and Mira-Titan datasets, generating posterior mean estimates 
and predicting at held-out cosmological settings. Across both datasets, DGP.FCO 
consistently performs well against competitive baselines, showcasing strong generalization and 
flexibility in complex, functional data regimes.

Several avenues for future research arise. Beyond modeling dark matter power 
spectra, DGP.FCO could be extended to hydrodynamical simulation outputs in cosmology, which 
often share similar functional structures and error correlations. More broadly, this approach may be 
applicable in other scientific domains involving structured functional outputs, especially those 
with spatial or temporal dependencies.

In addition, generating predictions over large cosmological grids opens the door to detailed 
sensitivity analyses, enabling researchers to disentangle parameter interactions and identify 
the most influential physical drivers of the spectra. From a modeling standpoint, further 
developments could include joint modeling of latent layers across cosmologies, and extensions 
to redshifts beyond $z=0$. Lastly, a hybrid framework combining DGP.FCO with Cosmic Emu, leveraging 
the strengths of each at different scales of $k$, may yield more accurate predictions.
